\thetitle

### 2020-12-28 paper/paper.tex
algorithm, near-Blackwell-optimality

Only if a policy is \(n\)-discount-optimal for all \(n < m\) it can be \(m\)-discount
optimal \citep{Puterman94,Veinott69}.
%
If a policy is \(\infty\)-discount-optimal then it is said to be Blackwell-optimal
\citep{Blackwell62}.
% weakening the argument!!!
That is, for Blackwell-optimal policies \(\polopt\) there exists a discount factor \(\gammaopt < 1\)
such that \(V_{\gamma}^{\polopt}(s) \geqslant V_{\gamma}^{\pol}(s)\) for all
\(\gamma \geqslant \gammaopt\) and under all policies \(\pol\)
\citep{Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning,Blackwell62}.
Informally that means there exists a discount factor \(\gamma <1\) which finds Blackwell-optimal
policies. However, (i) in more complex, i.e.\@ real world, MDPs this value can be arbitrary close to
\(1\) and (ii) the difference in state values may be very small, which (iii) due to the need of
state value function approximation likely causes errors when choosing among different actions in the
discounted framework.

% In the sequel we present the Laurent series expansion which not only links discounted reinforcement
% learning with average reward reinforcement learning, but also connects to \(n\)-discount-optimality.

% Furthermore, for the rest of the paper we occasionally write drop the \(V(s)\) and $V_{\gamma}$ instead of
% \(V^{\pol}(s)\) and \(V^{\pol}_{\gamma}\) when the policy is clear from the context.

The Bellman Equation defined as
\(V_{\gamma}^{\pol}(s) = \E_{\pol}[r + \gamma V_{\gamma}^{\pol}(s')]\) \citep[see
e.g.][p.70]{sutton1998introduction} provides a way to assess the state values using two
consecutively observed states \(s, s'\) and the observed reward \(r\) returned by the system. When
this formula is used as an update rule it provides an algorithm that converges the state values by
iteratively adapting the state value estimates.
%
There are four major issues with discounted RL, which results in the fact that it is inapplicable
for many problems of operations research.

\begin{enumerate}
\item Standard discounted RL can only infer suboptimal policies as the discount factor is strictly
  less than one, i.e.\@ \(\gamma < 1\).
\item It is very difficult to specify a desired balance between the short-term and long-term
  (average reward) rewards. This results from the fact that the long-term rewards are scaled
  exponentially, where the factor depends on the chosen \(\gamma\) value.
\item Episodic MDPs with an average reward per step that is non-zero cannot be solved, as the
  average reward is ignored in the terminal states.
\item The average reward is independently assessed for each state, even though it is the dominating
  part for all state values and thus is shared between more than one or, for unichain MDPs, even all
  states. This usually leads to an exponentially increased number of learning steps required for
  policies to converge.
\end{enumerate}

@@

Especially due to the fourth issue standard discounted RL is not well applicable in the operations
research domain. This mainly results from the fact that a continuous assessing of the state values
using costs or profit is often required, which leads to intractable long learning phases with high
exploration rate even for small sized problems. But as high exploration rates produce errors in the
state value estimations \citep{MillerVeinott1969} finding a well working parameterisation is
difficult. Average reward reinforcement learning overcomes these issues by separately assessing
addends of the state values.

@@

In the sequel we briefly present the needed formalism of average reward adjusted RL, where for a
more comprehensive version see \citep{schneckenreither2020average}. \MS[t]{todo}

\begin{definition}
  Due to \citet{schneckenreither2020average} the \textit{average reward adjusted (discounted) state
    value} $\X_{\gamma}^{\pol}(s)$ of a state $s$ under policy $\pol$ and with discount factor
  $0 \leqslant \gamma \leqslant 1$ is given as
  \(\X_{\gamma}^{\pol}(s) \defsym V^{\pol}(s) + e_{\gamma}^{\pol}(s)\).
\end{definition}
%
With the Laurent Series expensive this reformulates to
\(\X_{\gamma}^{\pol}(s) = V_{\gamma}^{\pol}(s) - \frac{\avgrew^{\pol}}{1-\gamma}\). Therefore,
\(X_{\gamma}^{\pol}(s)\) describes the discounted state value adjusted by the average reward term.
The Bellman Equation is given by
\(\X_{\gamma}^{\pol}(s) = \E_{\pol}[r_{t} + \gamma \X_{\gamma}^{\pol}(s_{t+1}) - \avgrew^{\pol}
\mid s_{t} = s]\). By turning this formula into an update rule it builds the foundation of our
algorithm.

% \subsection{Algorithm}
% \label{subsec:Algorithm}

\begin{algorithm}[t!]
  \begin{algorithmic}[1]
    \State{}Initialise state \(s_{0}\) and network parameters \(\netTarget, \netWorker\) randomly, set
    an exploration rate \(0 \leqslant \plearn \leqslant \pexp \leqslant 1\), exponential smoothing learning rates
    \(0 < \alpha, \gamma < 1\), and discount factors
    \(0 < \gamma_{0} < \gamma_{1} \leqslant 1\), where \(\gamma_{1} = 1\) is
    usually a good choice.
    \While{the stopping criterion is not fulfilled}
    \State{}\begin{minipage}[t]{0.9\textwidth} With probability \(\pexp\) choose a random action and
      probability \(1-\pexp\) one that fulfills
      \(\max_{a}\lexeq(\X^{\pol}_{\gamma_{1}}(s_{t},a; \netWorker{}),\X^\pol_ {\gamma_{0}}(s_{t},a; \netWorker{}))\). Let
      \(\isRandAct{t}\) indicate if the action was chosen randomly.
    \end{minipage}
    \State{}\begin{minipage}[t]{0.9\textwidth}
    Carry out action \(a_{t}\), observe reward \(r_{t}\) and resulting state \(s_{t+1}\).
    Store the experience \((s_{t}, \isRandAct{t}, a_{t}, r_{t}, s_{t+1})\) in the experience replay
    memory \(M\).
    \end{minipage}
    \If{a non-random action was chosen or \(\pexp > \plearn\)}
    \begin{align*}
      \avgrew^{\pol}  \gets (1- \alpha) \avgrew^{\pol} + \alpha [r_{t} +
      \max_{a}\X^\pol_{\gamma_1}(s_{t+1},a; \netWorker{}) - \X^\pol_{\gamma_1}(s_{t},a_{t}; \netWorker{})]
    \end{align*}
    \EndIf

@@

\paragraph{Average Reward Adjusted Deep RL Algorithm.} %
The model-free average reward adjusted RL algorithm is based on the tabular version of
\citet{schneckenreither2020average} and depicted in Algorithm~\ref{alg:near}.
%
%
%
\MS[t]{describe network initialisation method (Uniform, HeAtAl, etc.)}
%
\MS[t]{Extensions: overestimate rho, rho minimum, Multiple Agents (shared state,  shared rho?),
  init-phase (0.5 for expSmthRewRate, always adapt rho, same rho for all agents): exp smooth reward
  as rho, replay memory actions, n-step (describe replay memory), gradient clipping }
\MS[t]{Side info: Replay memory also stores (disallowed) filtered action indices for optimisation purposes and on
episodic tasks needs to store a bool indicating an episode end. Uses Int8 for state features values.  }

%
%
After initialising all values the agent enters the loop in which the first task is to choose an
action (step 3). In this action selection process we utilise an \(\epsilon\)-sensitive lexicographic
order \(a=(a_{1},\ldots,a_{n}) \lex (b_{1},\ldots,b_{n})=b\) defined as \(a \lex b\) if and only if
\(| a_{j} - b_{j} | \leqslant \epsilon\) for all \(j < i\) and \(|a_{i} - b_{i}| > \epsilon\). Note
that the resulting sets of actions may not be disjoint. Although this is an unusual order in
programming, taking the maximum as in our algorithm is straight-forward and thus cheap to compute.
%
The first action selection criteria maximises the average reward. In case of unichain and thus
ergodic MDPs only one scalar average reward value over all states is learned and thus the problem
simplified by abolishing the comparison in the first component. Then the set of actions which
maximise the bias values are selected. Finally, \(\Delta V^\pol_\gamma(s,a)\) is used to estimate
the slope of the error term \(e_{\gamma}^{\pol}(s,a)\), which can be either positive or negative.
Therefore, there is the need of a distinction of which actions to maximise according to the
discounted state value difference
\(\Delta V^{\pol}_{\gamma}(s,a) = V^\pol_{\gamma_1}(s,a) - V^\pol_{\gamma_0}(s,a)\). As the aim is
to collect rewards as soon as possible, a Blackwell-optimal reinforcement learning agents maximises
the error term, thus any actions converging to
\(\frac{\avgrew^{\pol}(s,a)}{1-\gamma_{0}} + \V^{\pol}(s,a)\) from above are preferred over those
that converge from below. Within these sets the action which maximises
\(\Delta V^{\pol}_{\gamma}(s,a)\) is selected. If multiple such actions the agent can choose
randomly among these.

As usual in reinforcement learning the equations are integrated by exponentially smoothing the
values utilising parameters \(\alpha, \beta, \gamma, \delta\).
% Further, as in this paper we concentrate on unichain MDPs the average reward
% \(\avgrew^{\pol}(s,a)\) can be learned as a scalar value \(\avgrew^{\pol}\), whereas the algorithm
% is supports multichain MDPs, cf. step \(5a\).
Step \(5a\) uses the Bellman equation for average reward reinforcement learning
(cf.~\citealt{Howard64, tadepalli1998model}) to find the average reward. Then step \(5b\), \(5c\)
and \(5d\) depict Equations (\ref{eq:constr2}) and (\ref{eq:constr3}) for \(n=0\), \(n=1\) and
\(n=2\). The need of learning values for \(n=2\) stems from the fact that the actual values
corresponding to the highest \(n\) are offset by vector \(u\).

% For \(n=0\) (step \(5b\)) the bias values are learned and further to ensure a unique solution of
% \(V^{\pol}\) step \(5c\) solves the constraints for \(n=1\). Thus, the first three assignments of
% step \(5\) correspond to the above constraint system for \(n=-1,0,1\). The last line of step \(5\)
% evaluates the deviation of \(W^{\pol}(s,a)\), which ought to be \(0\) once the system is correctly
% solved and if an action of the optimal policy \(\polopt\) was taken.
Step \(6\) infers the discounted state values using learning rate \(\gamma\) and two discount rates
\(0.5 \leqslant \gamma_{0} < \gamma_{1} < 1\) to estimate the state value difference
\(\Delta V^{\pol}_{\gamma}(s,a)\).

Steps \(7\) and \(8\) are crucial, but were not considered in average reward reinforcement learning
before. They feed back the deviations \(\Psi_{\cdot}(s,a)\) into the constraint problem creating a
recursive system of equations. Depending on the underlying MDP when this constraint is omitted one
likely observes steadily increasing deviations and therefore a wrong solution. This feedback loop
however is only activated in case a non-random action was chosen, as otherwise the deviations
\(\Psi_{\cdot}(s,a) > 0\) \citep{MillerVeinott1969}.

%
%
% Steps \(5d\) and \(5e\) calculate the deviations of actual values and expected values. In case the
% agent performs a non-random action
%
Recall that the algorithm evaluates the slope of \(e_{\gamma}^{\pol}(s)\) by learning the state
values using two different \(\gamma\) settings and discounted reinforcement learning. The slope of
\(e_{\gamma}^{\pol}(s)\) combined with the observation that for Blackwell-optimal policies
\(\polopt\) there exists a discount factor \(\gammaopt < 1\) such that
\(V_{\gamma}^{\polopt}(s) \geqslant V_{\gamma}^{\pol}(s)\) for all \(\gamma \geqslant \gammaopt\)
and under all policies \(\pol\)
\citep{Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning,Blackwell62}
provides the most selective criteria for our algorithm, which theoretically can find
Blackwell-optimal policies\footnote{This claim is only not fully proven yet. See the appendix for a
  first version of the proof. } (see Section~\ref{sec:methods} in Appendix).

@@

>>>>

Only if a policy is \(n\)-discount-optimal for all \(n < m\) it can be \(m\)-discount
optimal \citep{Puterman94,Veinott69}.
%
If a policy is \(-1\)-discount-optimal it is called gain-optimal, if it is \(0\)-discount-optimal it
is also called bias-optimal.
%
Furthermore, if a policy is \(\infty\)-discount-optimal then it is said to be Blackwell-optimal
\citep{Blackwell62}.
%
That is, for Blackwell-optimal policies \(\polopt\) there exists a discount factor \(\gammaopt < 1\)
such that \(V_{\gamma}^{\polopt}(s) \geqslant V_{\gamma}^{\pol}(s)\) for all
\(\gamma \geqslant \gammaopt\) and under all policies \(\pol\)
\citep{Mahadevan96_SensitiveDiscountOptimalityUnifyingDiscountedAndAverageRewardReinforcementLearning,Blackwell62}.
Informally that means there exists a discount factor \(\gamma <1\) which finds Blackwell-optimal
policies. However, (i) in more complex, i.e.\@ real world, MDPs this value can be arbitrary close to
\(1\) and (ii) the difference in state values may be very small, which (iii) due to the need of
state value function approximation likely causes errors when choosing among different actions in the
discounted framework. \citet{schneckenreither2020average} establishes a practically relevant
definition for near-Blackwell-optimality.

\begin{definition}
  If an algorithm infers for any MDP bias-optimal policies, and for a given MDP can in theory be
  configured to infer Blackwell-optimal policies, but in practise this ability is naturally limited
  due to the finite accuracy of floating-point representation of modern computer systems, it is said
  to be \emph{near-Blackwell-optimal}. An according to a near-Blackwell-optimal algorithm inferred
  Blackwell-optimal policy is called near-Blackwell-optimal.
\end{definition}

Note that standard discounted RL does not meet the requirements of near-Blackwell-optimality, as it
generally does not infer bias-optimal policies. Only by chance the resulting policy could be
bias-optimal.
%
% In the sequel we present the Laurent series expansion which not only links discounted reinforcement
% learning with average reward reinforcement learning, but also connects to \(n\)-discount-optimality.
%
% Furthermore, for the rest of the paper we occasionally write drop the \(V(s)\) and $V_{\gamma}$ instead of
% \(V^{\pol}(s)\) and \(V^{\pol}_{\gamma}\) when the policy is clear from the context.
%
The corresponding Bellman equation defined as
\(V_{\gamma}^{\pol}(s) = \E_{\pol}[r + \gamma V_{\gamma}^{\pol}(s')]\) \citep[see
e.g.][p.70]{sutton1998introduction} provides a way to assess the state values using two
consecutively observed states \(s, s'\) and the observed reward \(r\) returned by the system. When
this formula is used as an update rule it provides an algorithm that converges the state values by
iteratively adapting the state value estimates.
%
There are four major issues with discounted RL, which results in the fact that it is inapplicable
for many problems of operations research.

\begin{enumerate}
\item In general standard discounted RL can only infer suboptimal policies as the discount factor is
  strictly less than one, i.e.\@ \(\gamma < 1\).
\item It is very difficult to specify a desired balance between the short-term and long-term
  (average) rewards. This results from the fact that the long-term rewards are scaled exponentially,
  where the factor depends on the chosen \(\gamma\) value.
\item Episodic MDPs with an average reward per step that is non-zero cannot be solved correctly, as
  the average reward is ignored in the terminal states.
\item The average reward is independently assessed for each state, even though it is the dominating
  part for all state values and thus is shared between more than one or, for unichain MDPs, even all
  states. This usually leads to an exponentially increased number of learning steps required for
  policies to converge.
\end{enumerate}

@@

Especially due to the fourth issue standard discounted RL is not well applicable in the operations
research domain. This mainly results from the fact that a continuous assessing of the state values
using costs or profit is often required, which leads to intractable long learning phases with high
exploration rate even for small sized problems. But as high exploration rates produce errors in the
state value estimations \citep{MillerVeinott1969} finding a well working parameterisation is
difficult. Average reward reinforcement learning overcomes these issues by separately assessing the
subterms of the state values directly.

@@

Thus, average reward adjusted RL can be seen as a specialised version of average reward RL, where it
targets unichain MDPs only. In the sequel we briefly present the needed formalism of average reward
adjusted RL. For more comprehensive version we refer to \citet{schneckenreither2020average}.

\begin{definition}
  The \textit{average reward adjusted (discounted) state value} $\X_{\gamma}^{\pol}(s)$ of a state
  $s$ under policy $\pol$ and with discount factor $0 \leqslant \gamma \leqslant 1$ is given as
  \(\X_{\gamma}^{\pol}(s) \defsym V^{\pol}(s) + e_{\gamma}^{\pol}(s)\).
\end{definition}
%
With the Laurent Series expansion this reformulates to
\(\X_{\gamma}^{\pol}(s) = V_{\gamma}^{\pol}(s) - \frac{\avgrew^{\pol}}{1-\gamma}\). Therefore,
\(X_{\gamma}^{\pol}(s)\) describes the discounted state value adjusted by the average reward term.
The Bellman Equation is given by
\(\X_{\gamma}^{\pol}(s) = \E_{\pol}[r_{t} + \gamma \X_{\gamma}^{\pol}(s_{t+1}) - \avgrew^{\pol} \mid
s_{t} = s]\). The corresponding Bellman optimality equation then is given by
\(\X_{\gamma}^{\polopt}(s) = \max_{a} \E_{\polopt}[r_{t} + \gamma \X_{\gamma}^{\polopt}(s_{t+1}) -
\avgrew^{\pol} \mid s_{t} = s]\). By turning this formula into an update rule it builds the
foundation of our algorithm.

\begin{algorithm}[t!]
  \begin{algorithmic}[1]
    \State{}Initialise state \(s_{0}\) and network parameters \(\netTarget, \netWorker\) randomly, set
    an exploration rate \(0 \leqslant \plearn \leqslant \pexp \leqslant 1\), exponential smoothing learning rates
    \(0 < \alpha, \gamma < 1\), and discount factors
    \(0 < \gamma_{0} < \gamma_{1} \leqslant 1\), where \(\gamma_{1} = 1\) is
    usually a good choice.
    \While{the stopping criterion is not fulfilled}
    \State{}\begin{minipage}[t]{0.9\textwidth} With probability \(\pexp\) choose a random action and
      probability \(1-\pexp\) one that fulfills
      \(\max_{a}\lexeq(\X^{\pol}_{\gamma_{1}}(s_{t},a; \netWorker{}),\X^\pol_ {\gamma_{0}}(s_{t},a; \netWorker{}))\). Let
      \(\isRandAct{t}\) indicate if the action was chosen randomly.
    \end{minipage}
    \State{}\begin{minipage}[t]{0.9\textwidth}
    Carry out action \(a_{t}\), observe reward \(r_{t}\) and resulting state \(s_{t+1}\).
    Store the experience \((s_{t}, a_{t}, \isRandAct{t}, r_{t}, s_{t+1})\) in the experience replay
    memory \(M\).
    \end{minipage}
    \If{a non-random action was chosen or \(\pexp > \plearn\)}
    \begin{align*}
      \avgrew^{\pol}  \gets (1- \alpha) \avgrew^{\pol} + \alpha [r_{t} +
      \max_{a}\X^\pol_{\gamma_1}(s_{t+1},a; \netWorker{}) - \X^\pol_{\gamma_1}(s_{t},a_{t}; \netWorker{})]
    \end{align*}
    \EndIf

@@

\paragraph{Average Reward Adjusted Deep RL Algorithm.} %
The model-free average reward adjusted RL algorithm is based on the tabular version of
\citet{schneckenreither2020average} and depicted in Algorithm~\ref{alg:near}. The algorithm operates
on a target network \(\netTarget{}\) and a worker network \(\netWorker{}\) as in
\citet{mnih2015human}. % The target network reduces fluctuations in the state value estimates and
% therefore reduces the
%
In the action selection process a \(\epsilon\)-sensitive lexicographic order
\(a=(a_{1},\ldots,a_{n}) \lex (b_{1},\ldots,b_{n})=b\) is used. It is defined as \(a \lex b\) if and
only if \(| a_{j} - b_{j} | \leqslant \epsilon\) for all \(j < i\) and
\(|a_{i} - b_{i}| > \epsilon\). Note that the resulting sets of actions may not be disjoint, but
taking the maximum as required in our algorithm is straight-forward and thus cheap to compute. The
first action selection criteria selects the actions which maximise the bias values, the second the
error term.
%
Equation~\ref{eq:constr2} of the average reward RL constraints is used to estimate the average
reward based on the \MS[t]{reformulate avg RL constraints using s' etc.?} currently inferred state
values of two consecutive states and the returned reward. % As the average
% reward is equal among all states estimating the average reward is based on many observations.
This infers a solid average reward prediction of the current policy in few steps only and thus
quickly leads to better policies, as ones with higher gain are preferred.
\citet{schneckenreither2020average} and
\citet{Tadepalli98_ModelbasedAverageRewardReinforcementLearning} find that updating the average
reward by Equation~\ref{eq:constr2} is superior as compared to exponentially smoothing the actual
observed rewards. This makes sense as the later rather evaluates past policies.

The average reward adjusted state value, which is approximated based on the Bellman optimality
equation, is estimated twice. Once to infer the bias values with a very high discount factor
\(\gamma_{1}\), e.g.\@ \(\gamma_{1} = 1.0\), and a second time with a discount factor
\(\gamma_{0} < \gamma_{1}\). The later is used to more selectively choose actions in case more than
one action leads to bias-optimal policies. The predetermined discount value \(\gamma_{0}\) is a
measure for farsightedness. Lower values prefer to partially collect reward in fewer steps and
higher values are used to prefer actions for which the full bias is collected earlier.
%
The corresponding update values are saved to calculate the gradients based on the quadratic errors.
Every \(C\) steps the target network exponentially adapts the parameters of the worker network.
%
%
%
%
\MS[t]{describe network initialisation method (Uniform, HeAtAl, etc.)}
%
\MS[t]{Extensions: overestimate rho, rho minimum, Multiple Agents (shared state,  shared rho?),
  init-phase (0.5 for expSmthRewRate, always adapt rho, same rho for all agents): exp smooth reward
  as rho, replay memory actions, n-step (describe replay memory), gradient clipping }
\MS[t]{Side info: Replay memory also stores (disallowed) filtered action indices for optimisation purposes and on
episodic tasks needs to store a bool indicating an episode end. Uses Int8 for state features values.  }

@@

\MS[t]{Algorithm adaptions}
